\huge{Male-biased sexual selection, but not sexual dichromatism, predicts speciation

### 2018-10-20 Manuscript.Rmd
ReadMe Changes

---
title: \huge{Male-biased sexual selection, but not sexual dichromatism, predicts speciation
  in birds}
author:
- Justin G. Cally*\textsuperscript{§}, Devi Stuart-Fox\textsuperscript{§}, Luke Holman\textsuperscript{§}
  and Iliana Medina Guzmán\textsuperscript{§}
- ~
- '*\textit{justin.g.cally@gmail.com}'
bibliography: References/SS_Extinction.bib
output:
  pdf_document: default
  word_document: default
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage{hyperref}\hypersetup{colorlinks=true,
  linkcolor=blue, filecolor=blue, urlcolor=blue} \raggedbottom
csl: References/evolution.csl
subparagraph: yes
abstract: 'Sexual selection can shape phylogenetic diversity by altering speciation
  or extinction rates. The predicted effects on diversity are often contradictory;
  for example, sexual selection might promote speciation by creating reproductive
  isolation, or hinder it by selecting for traits that facilitate gene flow. Similarly,
  sexual selection might precipitate extinction by favouring competitive phenotypes
  with poor survival, or stave it off by helping populations shed deleterious mutations.
  Here, we investigate the relationship between sexual selection and rates of speciation
  and extinction in passerine birds. We use two measures of sexual selection --- sexual
  dichromatism (*n* = 5,812 species) and a multivariate measure of male-biased sexual
  selection (sexual dimorphism, polygyny and paternal care; *n* = 2,465) --- as well
  as three distinct measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$),
  and one measure of extinction ($\mu_{BAMM}$). We also tested whether the effects
  of sexual selection is contingent on ecological variables such as spatiotemporal
  temperature variation, primary productivity, and range size. We find that birds
  with smaller ranges have higher speciation rates; although the direction of causality
  is unclear. Moreover, our results show that male-biased sexual selection, but not
  sexual dichromatism, is an agent of adaptation and speciation. This finding suggests
  that sexual selection is an evolutionary force shaping patterns of diversity amongst
  songbirds, which has broad implications for ecology and conservation biology. \vspace{5mm}  \par\noindent \textbf{Keywords:} Speciation rate, Comparative analysis, Plumage dichromatism, Range size, Extinction rate.'
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Victoria, Australia.}

@@

Sexual selection results from competition for matings, fertilisations, or associated resources, and is typically stronger in males than females [@Fromhage_2016; @Janicke_2016; @Kokko_Jennions2008]. This evolutionary force has been proposed to facilitate speciation through the evolution of divergent signals associated with improved mating success [@Lande_1981; @Lande_1982]. For instance, in antbirds (Thamnophilidae), genera with complex songs and striking dichromatism are more species rich [@Seddon_2008]. Under this model of speciation, divergent mating signals in one sex (usually males) co-evolve with divergent preferences for those signals in the opposite sex, leading to behavioural reproductive isolation [@Safran_2013]. Alongside facilitating the evolution of divergent mating signals, sexual selection can increase the extent populations diverge by enlarging the available phenotypic space for mating signals. These predictions emerge as sexual selection may increase genetic diversity as a result of trade-offs between naturally- and sexually-selected traits or trade-offs under *intra*-locus sexual conflict between male and female trait expressions [@Lorch_2003; @Bonduriansky_2011; @Radwan_2016]. Additionally, sexual selection can have both positive and negative consequences for adaptation [*e.g.*, @Kokko_Brooks_2003; @Whitlock_Agrawal_2009; @Holman_Kokko_2013]. For example, sexual selection can elevate average fitness by helping to purge deleterious mutations [@Agrawal_2001; @Siller_2001] and fix beneficial ones [@Whitlock_2000], which should mitigate extinction risk and facilitate adaptation to environmental change. However, sexual selection might also promote extinction, by selecting for male traits that have detrimental pleiotropic effects on female traits [@Berger_2014; @Bonduriansky_Chenoweth_2009; @Pennell_2013; @Pischedda_2006], or selecting for male phenotypes with improved mating success but which harm population productivity, such as under-investing in parental care [@RN277].

Importantly, the effects of sexual selection on adaptation and speciation may depend on environmental conditions. Both theoretical and empirical studies suggest that under stressful/changing environments, sexual selection may have greater fitness benefits than under benign environments. Stressful environments strengthen the positive correlation between male mating success and female fitness, reducing the burden of negative pleiotropic effects [@Berger_2014; @Connallon_Hall_2016; @Long_2012; @Martinossi-Allibert_Savković_2017]. Conversely, benign or stable environments ensure consistent selection, preferentially eroding genetic variation at sexually concordant loci (*i.e.* loci where the optimum genotype is equivalent for both sexes). In these stable environments, genetic variation remains disproportionately at sexually antagonistic loci; causing displacement of male and female traits from their optimum. It follows from these predictions that populations with stronger sexual selection would have fitness benefits, allowing them to adapt to novel environments [*reviewed in* @Candolin_2008]. Additionally, sexual selection can reinforce locally adapted phenotypes through mate choice on phenotypes that effectively advertise quality in a specific ecological context [@Boughman_2002; @Maan_2011]. For instance, beak morphology is a trait that arose under natural selection in several taxa and is now a trait under sexual selection [a.k.a. *magic traits*; @Servedio_2011]. In Darwin's finches (*Geospiza fortis*) divergent beak morphology is an adaptation to local food availability that has been maintained through assortative mating [@Huber_2007]. Through these synergetic effects sexual selection and natural selection --- such as environmental change over space and time --- may lead to increased adaptation and speciation rates.

As a speciose and well-characterised group of organisms with remarkable and variable sexually-selected traits such as song and colourful plumage, birds are a popular focus for macroevolutionary studies of sexual selection and diversification [*e.g*, @Barraclough_1995; @Seddon_2008; @Huang_2014; @Seddon_2013; @Morrow_Speciation_2003]. A 2011 meta-analysis, covering 20 studies of birds and other taxa, found a small but significant positive association between sexual selection and speciation, with the average effect size in birds stronger than in mammals but weaker than in insects and fish [@Kraaijeveld_2011]. However, @Kraaijeveld_2011 found large variation in effect sizes between studies, likely because of differences in the metrics used as estimates of speciation and the strength of sexual selection. More recently, @Huang_2014 found no association between sexual dichromatism and speciation (*n* = 918 species) when using spectrophotometry measures on taxonomic specimens [@Armenta_2008] and tip-rate estimates from a molecular-only phylogeny [@Jetz_2012]. Similarly, @Cooney_2017 found no effect of sexual dichromatism on diversification across 1,306 pairs of species, using dichromatism scores from human observers.

Here, we investigate the association between sexual selection and diversification in birds while extending previous work in multiple ways. Firstly, many macroevolutionary studies estimate the strength of sexual selection using proxies such as sexual dimorphism or dichromatism, but these are imperfect correlates. We used a very large (*n* = 5,812) dataset of sexual dichromatism [estimated from illustrated drawings; @Dale_2015], as well as a smaller but still substantial dataset (*n* = 2,465) that gives a composite measure of male-biased sexual selection, capturing variation in sexual dimorphism, social polygyny and [lack of] paternal care. These datasets allow us to cover the great majority of bird genera, in contrast to previous work that has often focused on particular clades [*e.g.* @Seddon_2008]. Secondly, we use the latest methods for estimating speciation and extinction rates at the tips of phylogenetic trees, including BAMM (Bayesian Analysis of Macroevolutionary Mixtures) [*see*, @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017], as well as older but less controversial tip-rate statistics, such as diversification rate ($\lambda_{DR}$) and node density ($\lambda_{ND}$). We also conduct analyses that allow phylogenetic uncertainty in the avian tree [@Jetz_2012; *but see* @Rubolini_2015] to affect the precision of our estimates, rather than implicitly assuming that the single best-supported phylogenetic tree, or consensus tree, is correct. Thirdly, our analysis includes several potentially important environmental variables as predictors: range size, seasonal variation in temperature, spatial temperature variation (across a species’ range), long-term variation in temperature between now and the last inter-glacial (LIG) as well net primary productivity (NPP). By including these environmental variables in our models, we statistically account for their effects on variation in diversification rates and thereby improve our estimates of the effects of sexual selection. We also test whether the effects of sexual selection are contingent on the environment, as predicted by some theories [*e.g.* the theory that sexual selection is more likely to elevate adaptation in changeable relative to static environments; @Long_2012; @Connallon_Hall_2016].  \phantomsection

@@

We used a previously published measure of sexual dichromatism for 5,983 species of passerines [@Dale_2015]. Briefly, @Dale_2015 obtained sex-specific RGB (red, green and blue) values across six body patches (nape, crown, forhead, throat, upper breast and lower breast) from _Handbook of the Birds of the World_ (volumes 8–16) [@delhoyo2011]. The relative contribution of male and female RGB colour values were averaged across body patches and provide 'male-like' and 'female-like' plumage scores. Here we use the absolute difference between male and female plumage scores as an estimate of sexual dichromatism. Additionally, we used a more accurate measure of dichromatism corresponding to colour distance in avian colour space derived from spectral data [@Armenta_2008]. These measurements include variation in the ultraviolet and bird-visible range and --- unlike the RGB measures --- are sourced from taxonomic samples (as opposed to colour drawings). However, this spectrophotometry data is not as extensive as the RGB measures; with only 581 passerine species available for this analysis. While there is a correlation between these two measures, there is residual variation ($r = 0.79$; \href{https://justincally.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Figure S10}).  \phantomsection

@@

We obtained estimates of species range size using expert range maps [@birdlife_2017]. Because of taxonomic changes to 1,230 species in the _Birdlife_ database [@delhoyo_2016] we manually matched these taxa with the names used in the sexual dichromatism dataset [@Dale_2015]. From these distributions we obtained estimates of climatic conditions that each species is exposed to. That is, we took 1,000 random point samples in each range and extracted 19 present-day bioclimatic variables (representing a variety of biologically relevant annual trends in temperature and precipitation) with 30 seconds (~1 km^2^) spatial resolution [@Fick_2017]. From these values we obtained means and standard deviations for each species. Using the same spatial sampling, we extracted means and standard deviations of bioclimatic variables from the paleoclimate during the last interglacial (LIG; 120,000 - 140,000 years ago) [@Otto-Bliesner_2006]. Furthermore, to estimate variability in the energy available to species, we obtained the mean and standard deviation of net primary productivity (NPP) values between 2000 - 2015 across each species distribution. Estimates of NPP had 30 second resolution and were obtained through MODIS (Moderate Resolution Imaging Spectroradiometer) primary production products stage 3 (MOD17A3) [@Zhao_2005]. Using these data, which we provide as a potentially useful data resource (see \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM}), we generated five predictors of speciation associated with different patterns in environmental variability (see below).  \phantomsection

@@

Using the extracted environmental variables from each species range size we developed biologically meaningful predictors of environmental variation/stress relating to (*i*) seasonal climate variation, (*ii*) spatial climate variation and (*iii*) long-term climate variation. To obtain seasonal climate variation we used mean values of temperature seasonality (BIO4) for each range. To estimate levels of spatial environmental variation a species may endure we used the first principle component (PC1) from a PCA on standard deviations from all bioclimatic variables, excluding temperature (BIO4) and precipitation seasonality (BIO15). PC1 was heavily loaded towards bioclimatic variables relating to temperature, thus PC1 reflects the variation in temperature across a species' range (\href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Table S1}). Given that species range is a potentially informative predictor of speciation and extinction we controlled for the correlation between spatial variation and range size --- where larger ranges have larger variation in PC1 --- by taking the residuals of a fitted general additive model (GAM; \href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Figure S1}). To obtain long-term variation in climates for each species range we take the first principal component of the absolute difference in the bioclimatic variables between the LIG and current values. Similarly to spatial variation, the long-term climate variation is primarily loaded to temperature differences between the LIG and current climates (\href{https://justincally.github.io/SexualSelection_Speciation/#long_term_climate_variability_(lig_anomaly)}{Table S2, Figure S2}).The five predictors of environmental variability are not strongly correlated (\href{https://justincally.github.io/SexualSelection_Speciation/#correlations_between_environmental_predictors}{Figure S3}). Details and code to generate these predictors can be found within the \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM}.  \phantomsection

@@

To test the association between speciation/extinction and sexual selection, environmental variability and their interaction, we used phylogenetic least squares (PGLS) models in the ``nlme`` package [@nlme_2018]. We began by estimating the phylogenetic signal --- Pagel's $\lambda$ [@Pagel_1999] --- using the ``corPagel`` function in the ``ape`` package [@ape_2004] on the MCC tree. The estimate of $\lambda$ was then fixed for model selection, which was conducted on the same MCC tree derived from 2,500 draws of the posterior distribution [@Jetz_2012]. Model selection used either $\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$ or $\mu_{BAMM}$ tip-rate estimates from the MCC tree as the response variable. For $\lambda_{BAMM}$ and $\mu_{BAMM}$ we were also able to use model weights sourced from the inverse of the variance of tip-rates for a given species from the posterior distribution (*n* = 1,000). Model weights thus reflect the degree of precision to which each species tip-rate is measured in BAMM. Using model selection we only compared interaction terms between a measure of sexual selection (sexual dichromatism or male-biased sexual selection) and environmental measures. That is, we fixed the individual predictors of: dichromatism/male-biased sexual selection measures, log-transformed range size, seasonal temperature variation, spatial temperature variation, long-term temperature variation and NPP while comparing 32 models with different combinations of interactions (including none). Model selection was done in ``MuMIn`` using the ``dredge`` function [@MuMIn_2017]. Using the terms from the top-ranking model (lowest AICc), we ran the equivalent model for each of the 100 phylogenetic trees used to derive $\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$ and $\mu_{BAMM}$, using the unique response variables and phylogenetic tree correlation structure in each model. This method enabled us to present model estimates for an MCC tree alongside 100 trees from the posterior distribution of trees to account for phylogenetic uncertainty. This approach was repeated on three datasets corresponding to each measure of sexual selection: dichromatism derived from RGB values of images (*n* = 5,812); dichromatism from spectrophotometry (*n* = 581) and the multivariate measure of male-biased sexual selection (*n* = 2,465).

Finally, using data subset for species with measurements of male-biased sexual selection, we conducted a phylogenetic path analysis using the phylopath ``R`` package [@Wouter_2018]. The phylogenetic path analysis was used to assess causal paths *between* variables not able to be modelled within the univariate response of PGLS. That is, a phylogenetic path analysis allowed us to model relationships between the predictor variables used in our PGLS analysis as we anticipate environmental variability, sexual dichromatism/selection and range size to have effects on each other and not just on speciation rate. To minimise path complexity we use temperature seasonality (BIO4) as the single measure for environmental variability and use $\lambda_{DR}$ as the single measure of speciation. The phylogenetic path analysis used tip-rates and correlation structure from the MCC tree. Further details of the path analysis, including reasons for path directions, can be found within the \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM} along with all other analyses and the relevant ``R`` code to reproduce results.

>>>>

---
title: \huge{Male-biased sexual selection, but not sexual dichromatism, predicts speciation
  in birds}
author:
- Justin G. Cally*\textsuperscript{§}, Devi Stuart-Fox\textsuperscript{§}, Luke Holman\textsuperscript{§}
  and Iliana Medina Guzmán\textsuperscript{§}
- ~
- '*\textit{justin.g.cally@gmail.com}'
bibliography: References/SS_Extinction.bib
output:
  pdf_document: default
  word_document: default
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape} 
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} 
  \usepackage[hang, flushmargin]{footmisc} 
  \usepackage{hyperref}\hypersetup{colorlinks=true, hyperfootnotes=false, linkcolor=blue, filecolor=blue, urlcolor=blue}
  \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/evolution.csl
subparagraph: yes
abstract: 'Sexual selection can shape phylogenetic diversity by altering speciation
  or extinction rates. The predicted effects on diversity are often contradictory;
  for example, sexual selection might promote speciation by creating reproductive
  isolation, or hinder it by selecting for traits that facilitate gene flow. Similarly,
  sexual selection might precipitate extinction by favouring competitive phenotypes
  with poor survival, or stave it off by helping populations shed deleterious mutations.
  Here, we investigate the relationship between sexual selection and rates of speciation
  and extinction in passerine birds. We use two measures of sexual selection --- sexual
  dichromatism (*n* = 5,812 species) and a multivariate measure of male-biased sexual
  selection (sexual dimorphism, polygyny and paternal care; *n* = 2,465) --- as well
  as three distinct measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$),
  and one measure of extinction ($\mu_{BAMM}$). We also tested whether the effects
  of sexual selection is contingent on ecological variables such as spatiotemporal
  temperature variation, primary productivity, and range size. We find that birds
  with smaller ranges have higher speciation rates, although the direction of causality
  is unclear. Moreover, our results show that male-biased sexual selection, but not
  sexual dichromatism, is an agent of adaptation and speciation. This finding suggests
  that sexual selection is an evolutionary force shaping patterns of diversity amongst
  songbirds, which has broad implications for ecology and conservation biology. \vspace{5mm}  \par\noindent \textbf{Keywords:} Speciation rate, Comparative analysis, Plumage dichromatism, Range size, Extinction rate.'
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Victoria, Australia.}

@@

Sexual selection results from competition for matings, fertilisations, or associated resources, and is typically stronger in males than females [@Fromhage_2016; @Janicke_2016; @Kokko_Jennions2008]. This evolutionary force has been proposed to facilitate speciation through the evolution of divergent signals associated with improved mating success [@Lande_1981; @Lande_1982]. For instance, in antbirds (Thamnophilidae), genera with complex songs and striking dichromatism are more species rich [@Seddon_2008]. Under this model of speciation, divergent mating signals in one sex (usually males) co-evolve with divergent preferences for those signals in the opposite sex, leading to behavioural reproductive isolation [@Safran_2013]. Alongside facilitating the evolution of divergent mating signals, sexual selection can increase the extent populations diverge by enlarging the available phenotypic space for mating signals. These predictions emerge as sexual selection may increase genetic diversity as a result of trade-offs between naturally- and sexually-selected traits or trade-offs under *intra*-locus sexual conflict between male and female trait expressions [@Lorch_2003; @Bonduriansky_2011; @Radwan_2016]. Additionally, sexual selection can have both positive and negative consequences for adaptation [*e.g.*, @Kokko_Brooks_2003; @Whitlock_Agrawal_2009; @Holman_Kokko_2013]. For example, sexual selection can elevate average fitness by helping to purge deleterious mutations [@Agrawal_2001; @Siller_2001] and fix beneficial ones [@Whitlock_2000], which should mitigate extinction risk and facilitate adaptation to environmental change. However, sexual selection might also promote extinction by selecting for male traits that have detrimental pleiotropic effects on female traits [@Berger_2014; @Bonduriansky_Chenoweth_2009; @Pennell_2013; @Pischedda_2006], or selecting for male phenotypes with improved mating success but which harm population productivity, such as under-investing in parental care [@RN277].

Importantly, the effects of sexual selection on adaptation and speciation may depend on environmental conditions. Both theoretical and empirical studies suggest that under stressful/changing environments, sexual selection may have greater fitness benefits than under benign environments. Stressful environments strengthen the positive correlation between male mating success and female fitness, reducing the burden of negative pleiotropic effects [@Berger_2014; @Connallon_Hall_2016; @Long_2012; @Martinossi-Allibert_Savković_2017]. Conversely, benign or stable environments ensure consistent selection, preferentially eroding genetic variation at sexually concordant loci (*i.e.* loci where the optimum genotype is equivalent for both sexes). In these stable environments, genetic variation remains disproportionately at sexually antagonistic loci; causing displacement of male and female traits from their optimum. It follows from these predictions that populations with stronger sexual selection would have fitness benefits, allowing them to adapt to novel environments [*reviewed in* @Candolin_2008]. Additionally, sexual selection can reinforce local adaptations through mate choice on phenotypes that effectively advertise quality in a specific ecological context [@Boughman_2002; @Maan_2011]. For instance, beak morphology is a trait that arose under natural selection in several taxa and is now a trait under sexual selection [a.k.a. *magic traits*; @Servedio_2011]. In Darwin's finches (*Geospiza fortis*) divergent beak morphology is an adaptation to local food availability that has been maintained through assortative mating [@Huber_2007]. Through these synergetic effects sexual selection and natural selection --- such as environmental change over space and time --- may lead to increased adaptation and speciation rates.

As a speciose and well-characterised group of organisms with remarkable and variable sexually-selected traits such as song and colourful plumage, birds are a popular focus for macroevolutionary studies of sexual selection and diversification [*e.g.*, @Barraclough_1995; @Seddon_2008; @Huang_2014; @Seddon_2013; @Morrow_Speciation_2003]. A 2011 meta-analysis, covering 20 studies of birds and other taxa, found a small but significant positive association between sexual selection and speciation, with the average effect size in birds stronger than in mammals but weaker than in insects and fish [@Kraaijeveld_2011]. However, @Kraaijeveld_2011 found large variation in effect sizes between studies, likely because of differences in the metrics used as estimates of speciation and the strength of sexual selection. More recently, @Huang_2014 found no association between sexual dichromatism and speciation (*n* = 918 species) when using spectrophotometry measures on taxonomic specimens [@Armenta_2008] and tip-rate estimates from a molecular-only phylogeny [@Jetz_2012]. Similarly, @Cooney_2017 found no effect of sexual dichromatism on diversification across 1,306 pairs of species, using dichromatism scores from human observers.

Here, we investigate the association between sexual selection and diversification in birds while extending previous work in multiple ways. Firstly, many macroevolutionary studies estimate the strength of sexual selection using proxies such as sexual dimorphism or dichromatism, but these are imperfect correlates. We used a very large (*n* = 5,812) dataset of sexual dichromatism [estimated from illustrated drawings; @Dale_2015], as well as a smaller but still substantial dataset (*n* = 2,465) that gives a composite measure of male-biased sexual selection, capturing variation in sexual dimorphism, social polygyny and [lack of] paternal care [@Dale_2015]. These datasets allow us to cover the great majority of bird genera, in contrast to previous work that has often focused on particular clades [*e.g.* @Seddon_2008]. Secondly, we use the latest methods for estimating speciation and extinction rates at the tips of phylogenetic trees, including BAMM [Bayesian Analysis of Macroevolutionary Mixtures; *see*, @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017], as well as older but less controversial tip-rate statistics, such as diversification rate ($\lambda_{DR}$) and node density ($\lambda_{ND}$). We also conduct analyses that allow phylogenetic uncertainty in the avian tree [@Jetz_2012; *but see* @Rubolini_2015] to affect the precision of our estimates, rather than implicitly assuming that the single best-supported phylogenetic tree, or consensus tree, is correct. Thirdly, our analysis includes several potentially important environmental variables as predictors: range size, seasonal variation in temperature, spatial temperature variation (across a species’ range), long-term variation in temperature between now and the last inter-glacial (LIG) as well net primary productivity (NPP). By including these environmental variables in our models, we statistically account for their effects on variation in diversification rates and thereby improve our estimates of the effects of sexual selection. We also test whether the effects of sexual selection are contingent on the environment, as predicted by some theories [*e.g.* the theory that sexual selection is more likely to elevate adaptation in changeable relative to static environments; @Long_2012; @Connallon_Hall_2016].  \phantomsection

@@

We used a previously published measure of sexual dichromatism for 5,983 species of passerines [@Dale_2015]. Briefly, @Dale_2015 obtained sex-specific RGB (red, green and blue) values across six body patches (nape, crown, forhead, throat, upper breast and lower breast) from _Handbook of the Birds of the World_ [@delhoyo2011]. The relative contribution of male and female RGB colour values were averaged across body patches and provide 'male-like' and 'female-like' plumage scores. Here we use the absolute difference between male and female plumage scores as an estimate of sexual dichromatism. Additionally, we used a more accurate measure of dichromatism corresponding to colour distance in avian colour space derived from spectral data [@Armenta_2008]. These measurements include variation in the ultraviolet and bird-visible range and --- unlike the RGB measures --- are sourced from taxonomic samples (as opposed to colour drawings). However, this spectrophotometry data is not as extensive as the RGB measures; with only 581 passerine species available for this analysis. While there is a correlation between these two measures, there is residual variation ($r = 0.79$; \href{https://justincally.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Figure S10}).  \phantomsection

@@

We obtained estimates of species range size using expert range maps [@birdlife_2017]. Because of taxonomic changes to 1,230 species in the _Birdlife_ database [@delhoyo_2016] we manually matched these taxa with the names used in the sexual dichromatism dataset [@Dale_2015]. From these distributions we obtained estimates of climatic conditions that each species is exposed to. That is, we took 1,000 random point samples in each range and extracted 19 present-day bioclimatic variables (representing a variety of biologically relevant annual trends in temperature and precipitation) with 30 seconds (~1 km^2^) spatial resolution [@Fick_2017]. From these values we obtained means and standard deviations for each species. Using the same spatial sampling, we extracted means and standard deviations of bioclimatic variables from the paleoclimate during the last interglacial [LIG; 120,000 - 140,000 years ago; @Otto-Bliesner_2006]. Furthermore, to estimate variability in the energy available to species, we obtained the mean and standard deviation of net primary productivity (NPP) values between 2000 - 2015 across each species distribution. Estimates of NPP had 30 second resolution and were obtained through MODIS (Moderate Resolution Imaging Spectroradiometer) primary production products stage 3 [MOD17A3; @Zhao_2005]. Using these data, which we provide as a potentially useful data resource (see \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM}), we generated five predictors of speciation associated with different patterns in environmental variability (see below).  \phantomsection

@@

Using the extracted environmental variables from each species range size we developed biologically meaningful predictors of environmental variation/stress relating to (*i*) seasonal climate variation, (*ii*) spatial climate variation and (*iii*) long-term climate variation. To obtain seasonal climate variation we used mean values of temperature seasonality (BIO4) for each range. To estimate levels of spatial environmental variation a species may endure we used the first principle component (PC1) from a PCA on standard deviations from all bioclimatic variables, excluding temperature (BIO4) and precipitation seasonality (BIO15). PC1 was heavily loaded towards bioclimatic variables relating to temperature, thus PC1 reflects the variation in temperature across a species' range (\href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Table S1}). Given that species range is a potentially informative predictor of speciation and extinction, we controlled for the correlation between spatial variation and range size --- where larger ranges have larger variation in PC1 --- by taking the residuals of a fitted general additive model (GAM; \href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Figure S1}). To obtain long-term variation in climates for each species range we take the first principal component of the absolute difference in the bioclimatic variables between the LIG and current values. Similarly to spatial variation, the long-term climate variation is primarily loaded to temperature differences between the LIG and current climates (\href{https://justincally.github.io/SexualSelection_Speciation/#long_term_climate_variability_(lig_anomaly)}{Table S2, Figure S2}).The five predictors of environmental variability are not strongly correlated (\href{https://justincally.github.io/SexualSelection_Speciation/#correlations_between_environmental_predictors}{Figure S3}). Details and ``R`` code to generate these predictors can be found within the \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM}.  \phantomsection

@@

To test the association between speciation/extinction and sexual selection, environmental variability and their interaction, we used phylogenetic least squares (PGLS) models in the ``nlme`` package [@nlme_2018]. We began by estimating the phylogenetic signal --- Pagel's $\lambda$ [@Pagel_1999] --- using the ``corPagel`` function in the ``ape`` package [@ape_2004] on the MCC tree. The estimate of $\lambda$ was then fixed for model selection, which was conducted on the same MCC tree derived from 2,500 draws of the posterior distribution [@Jetz_2012]. Model selection used either $\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$ or $\mu_{BAMM}$ tip-rate estimates from the MCC tree as the response variable. For $\lambda_{BAMM}$ and $\mu_{BAMM}$ we were also able to use model weights sourced from the inverse of the variance of tip-rates for a given species from the posterior distribution (*n* = 1,000). Model weights thus reflect the degree of precision to which each species' tip-rate is measured in BAMM. Using model selection we only compared interaction terms between a measure of sexual selection (sexual dichromatism or male-biased sexual selection) and environmental measures. That is, we fixed the individual predictors of: dichromatism/male-biased sexual selection measures, log-transformed range size, seasonal temperature variation, spatial temperature variation, long-term temperature variation and NPP while comparing 32 models with different combinations of interactions (including none). Model selection was done in ``MuMIn`` using the ``dredge`` function [@MuMIn_2017]. Using the terms from the top-ranking model (lowest AICc), we ran the equivalent model for each of the 100 phylogenetic trees used to derive $\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$ and $\mu_{BAMM}$, using the unique response variables and phylogenetic tree correlation structure in each model. This method enabled us to present model estimates for an MCC tree alongside 100 trees from the posterior distribution of trees to account for phylogenetic uncertainty. This approach was repeated on three datasets corresponding to each measure of sexual selection: dichromatism derived from RGB values of images (*n* = 5,812); dichromatism from spectrophotometry (*n* = 581) and the multivariate measure of male-biased sexual selection (*n* = 2,465).

Finally, using data subset for species with measurements of male-biased sexual selection, we conducted a phylogenetic path analysis using the phylopath ``R`` package [@Wouter_2018]. The phylogenetic path analysis was used to assess causal paths *between* variables not able to be modelled within the univariate response of PGLS. That is, a phylogenetic path analysis allowed us to model relationships between the predictor variables used in our PGLS analysis as we anticipate environmental variability, sexual dichromatism/selection and range size to have effects on each other and not just on speciation rate. To minimise path complexity we used temperature seasonality (BIO4) as the single measure for environmental variability, $\lambda_{DR}$ as the single measure of speciation and the tip-rates from the MCC tree. Further details of the path analysis, including reasons for path directions, can be found within the \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM} along with all other analyses and the relevant ``R`` code to reproduce results.
